---
title: "Introduction to the BTWAR Package"
author:
- "Carlos Brás-Geraldes"
- "J. Leonel Rocha"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: BTWAR.bib
vignette: >
  %\VignetteIndexEntry{Introduction to the BTWAR Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 3.5,
  fig.align  = "center",
  out.width  = "90%"
)
```

# Introduction

The **BTWAR** package implements the Butterworth-Induced Autoregressive
modeling framework introduced in @btwar2026. The method derives
autoregressive (AR) coefficients directly from analog Butterworth filter
prototypes mapped into the discrete-time domain via the Matched Z-Transform
(MZT).

The central idea is to exploit the well-understood frequency-selectivity
properties of Butterworth filters --- maximally flat magnitude response,
monotonic roll-off, and a single tunable cutoff frequency --- and translate
them into a structured family of AR models. This establishes a principled
connection between **frequency-domain filter design** and **time-domain
autoregressive modeling**.

Unlike conventional AR modeling, where coefficients are estimated freely
from data, the BTW-AR model fixes the AR structure through two physically
meaningful hyperparameters: the filter order $N$ and the stopband
attenuation $A_s$. Only a single scalar gain $\alpha$ is estimated from
the data. This extreme parsimony has a direct practical consequence: the
fitted model admits a transparent frequency-domain interpretation that
purely data-driven approaches cannot provide.

The key advantages of the BTW-AR framework are:

- AR coefficients are fully determined by two interpretable parameters:
  filter order $N$ and stopband attenuation $A_s$.
- The cutoff frequency $f_c$ is derived analytically from $(N, A_s)$,
  providing a direct spectral interpretation of the selected model.
- Model order selection is performed via nested rolling-origin
  cross-validation, avoiding look-ahead bias.
- Scaling estimation supports three methods: least squares (LS), least
  absolute deviations (LAD), and Huber M-estimation, providing robustness
  options for non-Gaussian series.

# Methodology

## Butterworth Filter Prototype

An $N$-th order analog Butterworth lowpass filter with arbitrary gain
$H_0$ has transfer function

$$
H_a(s) = \frac{H_0}{\prod_{k=1}^{N}(s - s_k)},
$$

where the poles $s_k$ are located on a semicircle of radius $\Omega_c$
in the left half of the $s$-plane,

$$
s_k = \Omega_c \, e^{\,j\left(\frac{\pi}{2} + \frac{(2k-1)\pi}{2N}\right)},
\quad k = 1, \ldots, N,
$$

ensuring stability. 

In this framework, $H_0 = 1$ is adopted to preserve the spectral shape and establish a direct correspondence with the AR model in the $z$-domain.

## Matched Z-Transform

The MZT maps each analog pole $s_k$ to a discrete-time pole
$z_k = e^{s_k T}$, where $T = 1/f_s$ is the sampling period.
Since the Butterworth prototype is an all-pole filter with $H_0 = 1$,
there are no finite zeros, and the discrete-time transfer function
reduces to

$$
H(z) = \frac{1}{\prod_{k=1}^{N}\left(1 - z_k z^{-1}\right)},
\quad z_k = e^{s_k T},
$$

which preserves the all-pole structure required by an AR model.
Expanding the denominator polynomial via Viète's formulas gives

$$
\prod_{k=1}^{N}\left(1 - z_k z^{-1}\right)
= 1 + a_1 z^{-1} + a_2 z^{-2} + \cdots + a_N z^{-N},
$$

and the AR coefficients are obtained directly as $\phi_i = -a_i$,
so that the transfer function matches the standard AR form

$$
H(z) = \frac{1}{1 - \phi_1 z^{-1} - \cdots - \phi_N z^{-N}}.
$$

## Correspondence Between Filter Order and AR Order

The following result, established in @btwar2026, formalises the
equivalence between the number of digital poles and the number of lags in the autoregressive model.

---

**Proposition 2.**
*Let $H(z)$ be the transfer function of an autoregressive model $y(n)$
of order $p$. Then $H(z)$ has precisely $p = N$ lags, where $N$ is the
number of discrete-time poles $z_k = e^{s_k T}$, $k = 1, \ldots, N$.
In addition, the AR coefficients satisfy $\phi_i = -a_i$, where the
coefficients $a_i$ are given by Viète's formulas, for $i = 1, \ldots, p$.*

---

Proposition 2 has a direct practical consequence: the Butterworth filter
order $N$ simultaneously determines the autoregressive model order $p$.
There is therefore a one-to-one correspondence between the complexity of
the frequency-domain prototype and the temporal memory of the induced AR
model. In the example that follows, the PACF is used as a heuristic to
fix $p = 3$, which immediately sets $N = 3$ for the Butterworth design.

## The Butterworth-Induced Autoregressive Model

The principal theoretical result of @btwar2026 is the following theorem,
which formally defines the BTW-AR model and establishes the conditions
under which it is well-defined.

---

**Theorem 1** (Butterworth-Induced Autoregressive Model).
*Let $y(n)$ be an autoregressive model of order $p$, let $H_a(s)$ be the
Butterworth filter used to model its all-pole structure, and let
$\phi_i^{(BW)}$ denote the induced AR coefficients obtained via Viète's
formulas, for $i = 1, \ldots, N$, where $N$ is the minimum filter order
required to satisfy the stopband specification*

$$
N = \left\lceil \frac{1}{2}
    \frac{\log\!\left(10^{A/10} - 1\right)}
         {\log\!\left(f_{Nyquist}/f_c\right)}
\right\rceil,
$$

*and $A$ is the attenuation in decibels at the Nyquist frequency
$f_{Nyquist}$. If a scaling factor $\alpha$, estimated by minimising the
sum of squared residuals, satisfies $\varepsilon(n) \sim N(0,\sigma^2)$,
then the Butterworth-induced autoregressive model is defined by*

$$
y(n) = \alpha \sum_{i=1}^{N}
       \phi_i^{(BW)}\!\left(N,\, A,\, f_{Nyquist}\right)\, y(n-i)
       + \varepsilon(n).
$$

---

Three remarks are central to understanding this result.
First, the AR coefficients $\phi_i^{(BW)}$ are **entirely determined**
by the triplet $(N, A, f_{Nyquist})$ and are not estimated from the data;
only the scalar gain $\alpha$ is fitted.
Second, the filter order $N$ simultaneously controls the autoregressive
model order and the sharpness of the spectral roll-off.
Third, when the Gaussian residual assumption is violated, the least-squares
criterion for $\alpha$ can be replaced by a more general loss function

$$
J(\alpha) = \sum_{n} \rho\!\left(y(n) - \alpha\,\hat{y}(n)\right),
$$

accommodating LAD or Huber M-estimation without altering
the structural form of the model.

## Model Fitting Pipeline

Given a training series $\{y_t\}$, the fitting procedure is:

1. **Stationarity** --- apply successive differencing (guided by the
   Augmented Dickey-Fuller test) until stationarity is achieved.
2. **Centering** --- subtract the training mean $\bar{y}$.
3. **Cross-validation** --- search over a grid of $(A_s, N)$ pairs using
   nested rolling-origin CV to select the combination that minimises
   out-of-sample RMSE.
4. **Scaling** --- estimate a scalar $\alpha$ by regressing the observed
   series onto the one-step-ahead AR predictions using the chosen method
   (LS, LAD, or Huber).

# End-to-End Example

This section walks through a complete BTW-AR workflow using a simulated
AR(3) process. The example illustrates model fitting, spectral
interpretation via the Bode magnitude diagram, and the Z-plane pole
structure of the selected model.

## Setup

```{r setup-example, message = FALSE}
library(BTWAR)

set.seed(123)

fs <- 12000   # sampling frequency (Hz)
```

## Simulating Data

We simulate an AR(3) process with known coefficients and split it into
training (70%) and test (30%) sets using `simulate_ar_split()`.

```{r simulate}
phi_obs <- c(0.34, -0.22, 0.16)

sim_data <- simulate_ar_split(
  phi_real   = phi_obs,
  n          = 1000,
  burn       = 200,
  prop_train = 0.7
)

dados <- list(
  train = as.numeric(sim_data$train),
  test  = as.numeric(sim_data$test)
)
```

The simulated AR coefficients and their corresponding Z-plane poles are:

```{r true-params}
cat("True AR coefficients:\n")
print(phi_obs)

cat("\nTrue Z-plane poles:\n")
print(sim_data$poles)
```

## Exploratory Analysis

Following @btwar2026, the autoregressive order $p$ --- and consequently
the filter order $N$, in accordance with the correspondence established
by the MZT --- is first fixed using the partial autocorrelation function
(PACF) as a heuristic tool for identifying a suitable model order.
Once $N$ is fixed, the nested cross-validation procedure searches only
over the stopband attenuation $A$, reducing the hyperparameter search
to a one-dimensional grid.

```{r pacf, fig.cap = "PACF of the training series. Significant spikes at lags 1, 2, and 3 suggest AR order 3, which fixes the filter order N = 3."}
pacf(dados$train, main = "PACF – Training Series")
```

An inspection of the PACF suggests that an AR model with $p = 3$ lags
provides an adequate representation of the training series. This fixes
$N = 3$ prior to cross-validation.

## Fitting the BTW-AR Model

```{r fit, message = FALSE}
p <- 3

res <- btwar_fit(
  y_tr_raw = dados$train,
  y_te_raw = dados$test,
  fs       = fs,
  method   = "ls",
  N_vec    = p:p
)

cat("BTW-AR Performance:\n")
print(res$performance)
```

## Time-Series Predictions

The plots below show the one-step-ahead predictions of the BTW-AR model
on the training and test sets. The model tracks the observed signal
closely, capturing the dominant low-frequency dynamics of the process.

```{r plot-train, fig.cap = "Training set: observed series vs. BTW-AR predictions."}
plot(res, dataset = "train", fs = fs,
     colour_observed = "goldenrod",
     colour_btwar    = "blue",
     lwd_observed    = 1.2,
     lwd_btwar       = 1.2)
```

```{r plot-test, fig.cap = "Test set: observed series vs. BTW-AR predictions."}
plot(res, dataset = "test", fs = fs,
     colour_observed = "goldenrod",
     colour_btwar    = "blue",
     lwd_observed    = 1.2,
     lwd_btwar       = 1.2)
```

## Spectral Interpretation: Bode Magnitude Diagram

A distinctive feature of the BTW-AR framework is that the fitted model
admits a direct frequency-domain interpretation. The Bode magnitude
diagram below shows the frequency response of the Butterworth prototype
underlying the selected model.

```{r bode, fig.cap = "Butterworth magnitude response for the selected BTW-AR model.", fig.width = 10, fig.height = 5, out.width = "100%"}
plot_bode(res, fs = fs)
```

The diagram reveals two distinct spectral regions:

- **Passband** $[0, f_c]$ --- the model captures this band with a nearly
  flat gain, meaning that the low-frequency structure of the signal is
  represented with minimal distortion.
- **Transition and stopband** $[f_c, f_{Nyquist}]$ --- components in this
  region are progressively attenuated, reaching $A$ dB of suppression
  at the Nyquist frequency.

The cutoff frequency $f_c$ is therefore a direct, interpretable summary
of the model: it marks the spectral boundary between the signal dynamics
that the BTW-AR model captures and the higher-frequency variability that
it treats as noise. This frequency-domain transparency is not available
in conventional data-driven AR models, where the spectral implications
of the estimated coefficients are implicit and not immediately readable.

## Z-Plane Pole Structure

The Z-plane pole plot compares the poles of the fitted BTW-AR model with
those of the true data-generating process. Poles inside the unit circle
correspond to stable models.

```{r zpoles, fig.cap = "Z-plane poles: BTW-AR (selected) vs. true signal.", fig.height = 5}
df_true <- data.frame(
  Re = Re(sim_data$poles),
  Im = Im(sim_data$poles)
)

plot_zpoles(
  res,
  external_list = list(
    "True Signal Z-Poles" = df_true
  )
)
```

The pole locations of the BTW-AR model are entirely determined by the
Butterworth design parameters $(N, A)$ and are not estimated from the
data. Despite this structural constraint, the selected model places its
poles in a region of the Z-plane consistent with the dominant dynamics of
the true process, reflecting the effectiveness of the cross-validation
procedure in identifying a spectrally appropriate model.

# Robustness Options

The `method` argument of `btwar_fit()` controls how the scaling parameter
$\alpha$ is estimated from the data. Three options are available,
corresponding to different assumptions about the residual distribution.

| Method    | Loss function $\rho(e)$             | Recommended when                    |
|-----------|-------------------------------------|-------------------------------------|
| `"ls"`    | $e^2$                               | Gaussian residuals                  |
| `"lad"`   | $|e|$                               | Heavy-tailed or skewed noise        |
| `"huber"` | Quadratic-linear ($\delta = 1.345$) | Mild outliers present               |

Note that the AR coefficient structure is identical across all three
methods --- only the estimation of the scalar gain $\alpha$ changes.
This means that the spectral interpretation provided by the Bode diagram
remains valid regardless of the chosen method.

```{r robust, eval = FALSE}
fit_huber <- btwar_fit(
  y_tr_raw = dados$train,
  y_te_raw = dados$test,
  fs       = fs,
  method   = "huber",
  N_vec    = p:p
)
summary(fit_huber)
```

# Session Information

```{r session}
sessionInfo()
```

# References
