---
title: "Introduction to the BTWAR Package"
author:
  - "Carlos Brás-Geraldes"
  - "J. Leonel Rocha"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    highlight: tango
    fig_caption: true
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{hyperref}
  - \usepackage{xcolor}
  - \definecolor{codeblue}{RGB}{33, 102, 172}
  - \hypersetup{colorlinks=true, linkcolor=codeblue, urlcolor=codeblue, citecolor=codeblue}
  - \floatplacement{figure}{H}
vignette: >
  %\VignetteIndexEntry{Introduction to the BTWAR Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: BTWAR.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  out.width = "90%",
  fig.pos   = "H"
)
```

# Introduction

The **BTWAR** package implements the Butterworth-Induced Autoregressive
modeling framework introduced in @btwar2026. The method derives
autoregressive (AR) coefficients from analog Butterworth filter prototypes
mapped into the discrete-time domain via the Matched Z-Transform (MZT).

The key idea is to exploit the well-understood frequency-selectivity
properties of Butterworth filters --- maximally flat magnitude response,
monotonic roll-off, and a single tunable cutoff frequency --- and translate
them into a structured family of AR models. This establishes a principled
connection between **frequency-domain filter design** and **time-domain
autoregressive modeling**, with the following practical advantages:

- AR coefficients are fully determined by two interpretable parameters:
  filter order $N$ and stopband attenuation $A_s$.
- The cutoff frequency $f_c$ is optimised during fitting, providing a
  direct spectral interpretation of the selected model.
- Model order selection is performed via nested rolling-origin
  cross-validation, avoiding look-ahead bias.
- Scaling estimation supports three methods: least squares (LS), least
  absolute deviations (LAD), and Huber M-estimation, providing robustness
  options for non-Gaussian series.

# Methodology

## Butterworth Filter Prototype

An $N$-th order analog Butterworth lowpass filter has a squared magnitude
response

$$
|H(j\Omega)|^2 = \frac{1}{1 + \left(\Omega / \Omega_c\right)^{2N}}
$$

where $\Omega_c$ is the cutoff frequency. The poles of $H(s)$ lie on a
circle of radius $\Omega_c$ in the left half of the $s$-plane, uniformly
spaced in angle.

## Matched Z-Transform

The MZT maps each analog pole $s_k$ to a discrete-time pole
$z_k = e^{s_k T}$, where $T = 1/f_s$ is the sampling period.
The resulting AR polynomial

$$
A(z) = \prod_{k=1}^{N} \left(1 - z_k z^{-1}\right)
$$

defines the AR coefficients $\phi_1, \ldots, \phi_N$ via

$$
A(z) = 1 - \phi_1 z^{-1} - \cdots - \phi_N z^{-N}.
$$

## Model Fitting Pipeline

Given a training series $\{y_t\}$, the fitting procedure is:

1. **Stationarity** --- apply successive differencing (guided by the
   Augmented Dickey-Fuller test) until stationarity is achieved.
2. **Centering** --- subtract the training mean $\bar{y}$.
3. **Cross-validation** --- search over a grid of $(A_s, N)$ pairs using
   nested rolling-origin CV to select the combination that minimises
   out-of-sample RMSE.
4. **Scaling** --- estimate a scalar $\alpha$ by regressing the observed
   series onto the one-step-ahead AR predictions using the chosen method
   (LS, LAD, or Huber).

# Usage

## Installation

```r
# Install from CRAN
install.packages("BTWAR")

# Development version from GitHub
# remotes::install_github("cgeraldes/BTWAR")
```

## Simulated Example

We simulate an AR(2) series and partition it into training and test sets.

```{r simulate}
library(BTWAR)

set.seed(42)
n       <- 300
y       <- arima.sim(model = list(ar = c(0.7, -0.2)), n = n)
y_train <- y[seq_len(200)]
y_test  <- y[seq.int(201, 300)]
```

## Model Fitting

```{r fit}
fit <- btwar_fit(
  y_tr_raw = y_train,
  y_te_raw = y_test,
  method   = "ls"
)
```

The `summary()` method prints the selected model parameters and
performance metrics on both sets:

```{r summary}
summary(fit)
```

## Extracting Model Components

Individual components can be extracted using the standard S3 interface:

```{r extract}
# AR coefficients selected by cross-validation
coef(fit)

# First fitted values on the training set
head(fitted(fit))

# First training residuals
head(residuals(fit))
```

## Visualisation

The `plot()` method produces a \texttt{ggplot2} chart of observed versus
predicted values for either the training or test set.

```{r plot-test, fig.cap = "Test set: observed series (dark) vs. BTWAR predictions (blue)."}
plot(fit, dataset = "test", show_observed = TRUE)
```

An external benchmark series can be overlaid for direct comparison:

```{r plot-benchmark, eval = FALSE}
arima_fit  <- arima(y_train, order = c(1, 0, 0))
arima_pred <- predict(arima_fit, n.ahead = 100)$pred

plot(fit,
     dataset       = "test",
     show_observed = TRUE,
     external      = as.numeric(arima_pred),
     external_name = "ARIMA(1,0,0)")
```

## Spectral Inspection

The power spectrum of the series can be inspected with
`compute_spectrum()`:

```{r spectrum, fig.cap = "One-sided power spectrum of the simulated AR(2) series."}
sp <- compute_spectrum(as.numeric(y), fs = 1)

plot(sp$f, sp$P,
     type = "l", col = "#2166AC", lwd = 1.5,
     xlab = "Frequency (Hz)", ylab = "Power",
     main = "Power Spectrum")
```

# Robustness Options

The `method` argument of `btwar_fit()` controls how the scaling parameter
$\alpha$ is estimated. Table \ref{tab:methods} summarises the available
options.

\begin{table}[H]
\centering
\caption{Scaling estimation methods available in \texttt{btwar\_fit()}.}
\label{tab:methods}
\begin{tabular}{lll}
\toprule
\textbf{Method} & \textbf{Description} & \textbf{Recommended when} \\
\midrule
\texttt{"ls"}    & Ordinary least squares               & Gaussian residuals \\
\texttt{"lad"}   & Least absolute deviations            & Heavy-tailed or skewed noise \\
\texttt{"huber"} & Huber M-estimator ($\delta = 1.345$) & Mild outliers \\
\bottomrule
\end{tabular}
\end{table}

```{r robust, eval = FALSE}
fit_huber <- btwar_fit(
  y_tr_raw = y_train,
  y_te_raw = y_test,
  method   = "huber"
)
summary(fit_huber)
```

# Session Information

```{r session}
sessionInfo()
```

# References

