---
title: "Introduction to the BTWAR Package"
author:
- "Carlos Brás-Geraldes"
- "J. Leonel Rocha"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: BTWAR.bib
vignette: >
  %\VignetteIndexEntry{Introduction to the BTWAR Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  out.width = "90%"
)
```

# Introduction

The **BTWAR** package implements the Butterworth-Induced Autoregressive
modeling framework introduced in @btwar2026. The method derives
autoregressive (AR) coefficients from analog Butterworth filter prototypes
mapped into the discrete-time domain via the Matched Z-Transform (MZT).

The key idea is to exploit the well-understood frequency-selectivity
properties of Butterworth filters --- maximally flat magnitude response,
monotonic roll-off, and a single tunable cutoff frequency --- and translate
them into a structured family of AR models. This establishes a principled
connection between **frequency-domain filter design** and **time-domain
autoregressive modeling**, with the following practical advantages:

- AR coefficients are fully determined by two interpretable parameters:
  filter order $N$ and stopband attenuation $A_s$.
- The cutoff frequency $f_c$ is optimised during fitting, providing a
  direct spectral interpretation of the selected model.
- Model order selection is performed via nested rolling-origin
  cross-validation, avoiding look-ahead bias.
- Scaling estimation supports three methods: least squares (LS), least
  absolute deviations (LAD), and Huber M-estimation, providing robustness
  options for non-Gaussian series.

# Methodology

## Butterworth Filter Prototype

An $N$-th order analog Butterworth lowpass filter has a squared magnitude
response

$$
|H(j\Omega)|^2 = \frac{1}{1 + \left(\Omega / \Omega_c\right)^{2N}}
$$

where $\Omega_c$ is the cutoff frequency. The poles of $H(s)$ lie on a
circle of radius $\Omega_c$ in the left half of the $s$-plane, uniformly
spaced in angle.

## Matched Z-Transform

The MZT maps each analog pole $s_k$ to a discrete-time pole
$z_k = e^{s_k T}$, where $T = 1/f_s$ is the sampling period.
The resulting AR polynomial

$$
A(z) = \prod_{k=1}^{N} \left(1 - z_k z^{-1}\right)
$$

defines the AR coefficients $\phi_1, \ldots, \phi_N$ via

$$
A(z) = 1 - \phi_1 z^{-1} - \cdots - \phi_N z^{-N}.
$$

## Model Fitting Pipeline

Given a training series $\{y_t\}$, the fitting procedure is:

1. **Stationarity** --- apply successive differencing (guided by the
   Augmented Dickey-Fuller test) until stationarity is achieved.
2. **Centering** --- subtract the training mean $\bar{y}$.
3. **Cross-validation** --- search over a grid of $(A_s, N)$ pairs using
   nested rolling-origin CV to select the combination that minimises
   out-of-sample RMSE.
4. **Scaling** --- estimate a scalar $\alpha$ by regressing the observed
   series onto the one-step-ahead AR predictions using the chosen method
   (LS, LAD, or Huber).

# End-to-End Example

This section walks through a complete BTWAR workflow using a simulated
AR(3) process, including model fitting, comparison with ARIMA, and the
full suite of diagnostic plots.

## Setup

```{r setup-example, message = FALSE}
library(BTWAR)
library(forecast)

set.seed(123)

fs <- 12000   # sampling frequency (Hz)
```

## Simulating Data

We simulate an AR(3) process with known coefficients and split it into
training (70%) and test (30%) sets using `simulate_ar_split()`.

```{r simulate}
phi_true <- c(0.34, -0.22, 0.16)

sim_data <- simulate_ar_split(
  phi_real   = phi_true,
  n          = 1000,
  burn       = 200,
  prop_train = 0.7
)

dados <- list(
  treino = as.numeric(sim_data$train),
  teste  = as.numeric(sim_data$test)
)
```

The true AR coefficients and their corresponding Z-plane poles are:

```{r true-params}
cat("True AR coefficients:\n")
print(phi_true)

cat("\nTrue AR poles:\n")
print(sim_data$poles)
```

## Exploratory Analysis

A partial autocorrelation plot confirms the AR order before fitting:

```{r pacf, fig.cap = "PACF of the training series, suggesting AR order 3."}
pacf(dados$treino, main = "PACF – Training Series")
```

## Fitting BTWAR

```{r fit, message = FALSE}
p=3

res <- btwar_fit(
  y_tr_raw = dados$treino,
  y_te_raw = dados$teste,
  fs       = fs,
  method   = "ls",
  N_vec    = p:p
)

cat("BTWAR Performance:\n")
print(res$performance)
```

## Comparison with ARIMA

We fit an ARIMA(3,0,0) on the same centered training series and compute
one-step-ahead fitted values aligned with the BTWAR prediction window.

```{r arima}
fit_arima <- forecast::Arima(
  res$train$y_real,
  order = c(p, 0, 0)
)

arima_fitted_tr <- tail(
  as.numeric(fitted(fit_arima)),
  length(res$train$yhat)
)

arima_vals_te <- tail(
  as.numeric(fitted(forecast::Arima(dados$teste - res$mu, model = fit_arima))),
  length(res$test$yhat)
)

rmse_arima_te <- sqrt(mean((res$test$y_real - arima_vals_te)^2))

cat("RMSE Comparison:\n")
cat("  BTWAR Test RMSE :", res$performance$rmse_test, "\n")
cat("  ARIMA Test RMSE :", rmse_arima_te, "\n")
```

## Time-Series Plots

### BTWAR Predictions

```{r plot-train-btwar, fig.cap = "Training set: observed series vs. BTWAR predictions."}
plot(res, dataset = "train", fs = fs,
     colour_observed = "goldenrod",
     colour_btwar    = "blue",
     lwd_observed    = 1.2,
     lwd_btwar       = 1.2)
```

```{r plot-test-btwar, fig.cap = "Test set: observed series vs. BTWAR predictions."}
plot(res, dataset = "test", fs = fs,
     colour_observed = "goldenrod",
     colour_btwar    = "blue",
     lwd_observed    = 1.2,
     lwd_btwar       = 1.2)
```

### BTWAR vs. ARIMA

```{r plot-train-arima, fig.cap = "Training set: BTWAR vs. ARIMA one-step-ahead predictions."}
plot(res, dataset = "train", fs = fs,
     colour_observed = "goldenrod",
     colour_btwar    = "blue",
     colour_external = "red",
     lwd_observed    = 1.2,
     lwd_btwar       = 1.2,
     lwd_external    = 1.2,
     external        = arima_fitted_tr,
     external_name   = "ARIMA",
     layer_order     = c("Observed", "BTWAR", "ARIMA"))
```

```{r plot-test-arima, fig.cap = "Test set: BTWAR vs. ARIMA one-step-ahead predictions."}
plot(res, dataset = "test", fs = fs,
     colour_observed = "goldenrod",
     colour_btwar    = "blue",
     colour_external = "red",
     lwd_observed    = 1.2,
     lwd_btwar       = 1.2,
     lwd_external    = 1.2,
     external        = arima_vals_te,
     external_name   = "ARIMA",
     layer_order     = c("Observed", "ARIMA", "BTWAR"))
```

## Z-Plane Pole Plot

The Z-plane poles of all three models --- the true data-generating process,
BTWAR, and ARIMA --- are displayed together for direct comparison.
Poles inside the unit circle correspond to stable models.

```{r zpoles, fig.cap = "Z-plane poles: BTW–AR (selected), true signal, and ARIMA.", fig.height = 5}
# ARIMA poles
phi_arima <- coef(fit_arima)[grep("^ar", names(coef(fit_arima)))]
df_arima  <- data.frame(
  Re = Re(poles_AR(phi_arima)),
  Im = Im(poles_AR(phi_arima))
)

# True signal poles
df_true <- data.frame(
  Re = Re(sim_data$poles),
  Im = Im(sim_data$poles)
)

plot_zpoles(
  res,
  external_list = list(
    "Observed Signal Z-Poles" = df_true,
    "ARIMA Z-Poles"           = df_arima
  )
)
```

## Bode Magnitude Diagram

The Bode magnitude diagram shows the frequency response of the Butterworth
filter underlying the selected BTWAR model, with reference lines at the
cutoff and Nyquist frequencies.

```{r bode, fig.cap = "Butterworth magnitude response for the selected BTWAR model.", fig.height = 4}
plot_bode(res, fs = fs)
```

# Robustness Options

The `method` argument of `btwar_fit()` controls how the scaling parameter
$\alpha$ is estimated. The table below summarises the available options.

| Method | Description | Recommended when |
|--------|-------------|------------------|
| `"ls"` | Ordinary least squares | Gaussian residuals |
| `"lad"` | Least absolute deviations | Heavy-tailed or skewed noise |
| `"huber"` | Huber M-estimator ($\delta = 1.345$) | Mild outliers |

```{r robust, eval = FALSE}
fit_huber <- btwar_fit(
  y_tr_raw = dados$treino,
  y_te_raw = dados$teste,
  fs       = fs,
  method   = "huber",
  N_vec    = p:p
)
summary(fit_huber)
```

# Session Information

```{r session}
sessionInfo()
```

# References
